---
title: "Roadway Capacity"
bibliography: CF.bibtex
---
***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE, warning = FALSE, message = FALSE, error = FALSE}
library(Basic)
```


__Flow and Capacity__

Throughout our discussion of __cartools__, we have intentionally avoided discussing roadway capacity, a topic that has received much attention. In fact, we have not defined traffic flow as yet. The flow is approximated as average flow time traffic density [@deganzo:1997]:

* $q \approx \bar{u} * k$

where $q$ is treated as a constant value, $\bar{u}$ as a weighted average by vehicle class and $k$ as a constant value. Appearances are deceiving. The relationship is more complicated than it appears.  Let us  assume all vehicles are of the same class, say passenger cars and $k$ is a constant. The flow $q$ is not a constant.  It is a random variable that we denote as $Q$. thus $Q = q = \bar{u} * k$ where $Q$ is an expected value.

The definition of capacity that has been refined over the years. @banks:1998 simply defines capacity as:

* $Q^* = 1/\bar{h}_{min}$

where $Q^*$ is the maximum value of $Q$ and $\bar{h}_{min}$ is the minimum average headway. This estimate, which is applied to highway and rail systems, is appealing owing to its simplicity. @LiliBook defines the _HCM_ capacity  as a "maximum obtainable flow rate."  The $Q^*$ relationship satisfies this definition. Both authors recognize the difficulties of estimating and assigning capacity owing to the volatility of the field data. 


__Maximize Traffic Throughput__

We estimate $Q^*$  as a maximization problem with an objection function to  "maximize throughput." Thus, we rewrite out  $Q(t)  = k * U(t)$ model where $U(t) = U(k, u, \sigma, h_{safe}, t)$ as:


* maximize  $Q^*  = k^* * \bar{u}(k^*)$.

A major strength of the __Brownian bridge model of speed__, a stochastic model $U(k,t)$, is its ability to explain the role of traffic noise and role that traffic density $k$ plays in forecasting the probability of being in a congested state, $\pi(k).$ Since the $U(k,t)$ model is derived from a controlled experiment where $k$ is a fixed value, traffic flow, a most important measure of highway performance, is simply defined as a _parsimonius, stochastic model_:

* $Q(t)  = k * U(k, u, \sigma, h_{safe}, t)$.

To draw attention to the significance of the model in explaining traffic flow with $k$, we rewrite as a time-series model:

* $Q(t)  = k * U(t)$

where $U(t) = U(k, u, \sigma, h_{safe}, t)$. The model tacitly incorporates the roadway geometry (a bottleneck) and driver behavior as reflected by $h_{safe}$. Here is a reminder that the process is volatile and chancy:


```{r uk_sdkplot, message = FALSE, fig.height = 4, fig.width = 4, echo = FALSE, warning=FALSE}
par(mfrow = c(1,1), pty = "s")

df <- uk_sdk(QUKtdayX,TRUE)
```

```{r logisticplots, message = FALSE, fig.height =4, fig.width = 4, echo = FALSE, warning=FALSE}
par(mfrow = c(1,1), pty = "s")
df <- logisticmodel(QUKtdayX,TRUE)
```

__Optimum Solution__

The solution seems  straightforward enough. Select $k^*$. The following graph uses this information to aid in the decision-making. "Success" refers to the probability that the roadway is operating in a free-flow state, $P(X = 0) = 1 - \pi$.

```{r capacityplot, message = FALSE, fig.height = 4, fig.width = 4, echo = FALSE, warning=FALSE}
par(mfrow = c(1,1))
capacityplot(QUKtdayX)
```

The "maximum throughput" is  $Q^*$ = 2400 vph  and success probability is $P(X = 0)$ = 0.13 at  $k^*$ = 40 vph.  The HCM specifies $k^*$ as 45 vpm. According to our results, $Q$ = 2340 vph and $P(X = 0)$ = 0.6.

The importance of  $k$ is clearly  evident. For $k > k^*$, the $Q \approx 1800$ vph and $P(X = 0) \rightarrow 0$ performance measures suffer. Good performance is achieved for $k \leq k^*$ = 40 vph.

So, what is correct value of capacity? The best answer is "It depends." Consider the _HCM_ guidelines give a range of capacity values from 2250 to 2400 vph. The vale depends on the free-mean speed associated with the geometric design.  [@Laflamme2018a] give a range from 2400 to 2700 vph. Their approach uses an extreme-value value estimation approach and the analysis of maximum daily flow data including censored data. This estimation method helps explain why their capacity range is greater than the _HCM_. 


***

__References__
