---
title: "Tools For UNDERSTANDING Highway Performance (TFUHP)"
bibliography: "CF.bibtex"
output: html_document
---


```{r, include = FALSE, warning = FALSE, message = FALSE, error = FALSE}
library(shiny)
library(RJSONIO)
library(knitr)
library(tinytex)
```


The material presented in __TFUHP__ is developed from various sources, including full scale testing, transportation textbooks, research classified as _traffic flow theory_, and probability models. The following is an extremely small sample of the resources used in this development:

* Full Scale Testing: @1367-2630-10-3-033001

* Textbooks: @banks:1998, @deganzo:1997, @LiliBook, @trieber

* Research: @pjo2017 

* Probability Modeling: @iacus, @sde 

The abstract given in the @pjo2017 paper gives a good idea of the overall theme of __TFUHP__. Mathematical tools are used to _predict_ the probabilty of traffic breakdown, or stated another way, the likelihood of an onset of a congestion event. 

To reliably predict breakdown is not an easy problem. The root cause of traffic breakdown must be _identified and understood_ before reliable predictions can be made. 

In this paper, the number of vehicles measured as a traffic density (vehicles per unit length) is used as an explanatory variable in lieu of traffic flow (number vehicles passing a point per unit of time). Traffic is a highly volatile process, thus probability plays a vital role. Thus, both (1) traffic density and (2) traffic noise (volatily) are used to explain breakdown. 


```{r label, out.width = "100%", fig.cap = "Footnote: I haven't figured out how to print the full six-page document.", echo = FALSE}
knitr::include_graphics("IEEE_MT-ITS.pdf", auto_pdf = getOption("knitr.graphics.auto_pdf", TRUE))
```

[https://pjossen.shinyapps.io/headway/](https://pjossen.shinyapps.io/headway/)

## References


