---
title: "The Future"
bibliography: CF.bibtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE, warning = FALSE, message = FALSE, error = FALSE}
library(Basic)
```


__Intelligent Transportion Systems__ 

As we carefully review the outcomes of the simulations derived from various deterministic and stochastic models, we must conclude that self-optimizing drivers are working against their own self interest of moving forward with no or minimum delay. Contrast the assumptions of the __Zipper Merge__ and __Traffic Breakdown Part 6__ simulations. 

The __Zipper Merge__ simulation assumes that drivers act in concert to align themselves before merging. The analysis suggests that drivers can successful merge at any speed, suggesting that high speed merges can take place.  There is no empirical evidence to prove this contention other than observing drivers on a  race track.

__Traffic Breakdown Part 6__ simulations assume that drivers act individually. Drivers, who have no incentive to give their counterparts room to maneuver, promote delay. This kind of self-serving behavior is not new to traffic operations. It is known as _Braess's Paradox_ [@braess]. It has been shown both with mathematical modeling and with empirical evidence that closing certain roads in a roadway network can improve service.   For example, @Roughgarden:2002 reported that total delay using convex programming for a simple idealized network would be one-third greater for a "non-optimum" network than for an "optimum" network. 

Some people may claim that counter intuitive methods as suggested by the _Braess's Paradox_ approach has  no application to traffic breakdown at a bottleneck or at a lane drop.  It is far-fetched notion to assume the geometric layout of an interstate freeway interchange, say,  can be improved in any meaningful or practical way.  Secondly, training drivers to change their habits is social engineering  requiring unobtainable behavioral changes. For example, drivers are taught that the right and left freeway lanes are reserved for slower and higher speed travel, respectively. Take any freeway of your choosing and  observe that this teaching is not enforced by law enforcement and  ignored by the driving public. Tailgating is common practice. Thirdly and finally, critics can claim that the deterministic  __Zipper Merge__ model disregards traffic volatility, therefore it ignores a most fundamental premise of the work presented here in the __cartools__ package.

These are formidable criticisms. However, we have not considered controlling traffic noise while leaving  individual driver  behavior alone. How can this be done? An answer may be to apply  _Intelligent Transportion Systems_ (_ITS_) in a _smart city environment_. Consider this  proof of concept:

1. Collect real-time data. Data collected with lidar and radar equipment placed roadside or collected on-board a vehicle with a GPS system. Measures, including location $x_{vehicle}(t)$ and speed $u_{vehicle}(t)$  data will be transmitted using wireless technology and stored on a central controller computer.

2. Develop a __brkzipper__ forecast algorithm. The algorithm will be created along the lines of the __Zipper Merge__ and __Traffic Breakdown Part 6__ models. In lieu of simulating zipper and traffic breakdown merges, __brktrials__ algorithm will be designed to forecast $\hat{x}_{vehicle}(t)$ locations and $\hat{u}_{vehicle}(t)$ speeds at future times of $\tau$. 

3. Transmit the forecasts back to the vehicles. To be effective, drivers will relinquish control of their vehicles and an on-board control will be used. The on-board systems are contemplated to be autonomous units similar in operation to autonomous cruise control (ACC) devices used on contemporary vehicles sold today.

__Capacity__

Throughout our discussion of __cartools__, we have intentionally avoided discussing roadway capacity, a topic that has received much attention. So, what is correct value of capacity? The best answer is "It depends." Consider the _HCM_ guidelines give a range of capacity values from 2250 to 2400 vph on the geometric design associated with the free-mean speed.  [@Laflamme2018a] give a range from 2400 to 2700 vph. Their approach uses an extreme-value value estimation approach and the analysis of maximum daily flow data including censored data. The estimation method helps explain why their capacity range is greater than the _HCM_. 

Using these ranges for comparison, let's look at the estimation of capacity using the results obtained with __cartools.__ Before doing so, consider the definition of capacity that has been refined over the years. @LiliBook defines the _HCM_ capacity  as a "maximum obtainable flow rate."  She recognizes the difficulties of estimating and assigning capacity owing to the volatility of the field data. Using this definition, we will solve the problem by treating it as a maximization problem with an objection function to  "maximize throughput." Thus, we rewrite out  $Q(t)  = k * U(t)$ model where $U(t) = U(k, u, \sigma, h_{safe}, t)$ as:


* maximize  $Q^*  = k^* * \bar{u}(k^*)$.

The solution seems  straightforward enough. Select $k^*$. However, we must remind ourselves that we are dealing with volatile field data. As a reminder, the following two plots are reproduced.


```{r uk_sdkplot, message = FALSE, fig.height = 4, fig.width = 4, echo = FALSE, warning=FALSE}
par(mfrow = c(1,1), pty = "s")
df <- uk_sdk(QUKtdayX,TRUE)
```

```{r logisticplots, message = FALSE, fig.height = 4, fig.width = 4, echo = FALSE, warning=FALSE}
par(mfrow = c(1,1), pty = "s")
df <- logisticmodel(QUKtdayX,TRUE)
```

The following graph uses this information to aid in the decision-making. "Success" refers to the probability that the roadway is operating in a free-flow state, $P(X = 0) = 1 - \pi$.

```{r capacityplot, message = FALSE, fig.height = 4, fig.width = 4, echo = FALSE, warning=FALSE}
par(mfrow = c(1,1))
capacityplot(QUKtdayX)
```

The "maximum throughput" is  $Q^*$ = 2400 vph  and $P(X = 0)$ = 0.13 at  $k^*$ = 40 vph.  The importance of  $k^*$ is clearly  evident. For example at $k = 45$, $P(X = 0)$ = 0.6 and $Q$ = 2340 vph. At $k > k^*$, the $Q$ and $P(X = 0)$ performance measures suffer.

__Smart Cities and the Internet of Things__

Controlling traffic noise by using _ITS_ in a _smart city environment_, as suggested above, is expected to boost performance even further than presented here. The key questions are: "(1) By how much, (2) who is interested, and (3) who can supply the answers?" 

Question (1), the $\$1B$ question, requires further research to provide a good answer. If the  __cartools__  package offers nothing else, a reader must come away knowing that collecting good data  from controlled experiments and the good field data are essential. 

Once these data are obtained meaningful analyses can be conducted. Of course, even when objective is focused on  the understanding the effects of traffic noise and controlling it, an interdisciplinary approach is necessary. This does not mean establishing an academic, interdisciplinary team of computer scientists, engineers and statisticians. @albino state that there is no one formula for success but a clear need to engage policy makers from local communities.  

The cities of Chattanooga, TN and Syracuse, NY, have  a good working relationships and a dialog with University of Tennessee and Syracuse University that can be tapped. For example, the SU College of Engineering and Computer Science and its newly established  Invent@SU inventor accelerators have engaged undergraduates in various projects across disciplines and colleges that engage them with local communities in many cities across the country.  

Since _smart city_ initiatives use Big Data and often imply grand scale projects, it does not mean that a $\$1B$ in seed money is required to get it started. A handful of faculty, who are annually engaged in senior projects,  is necessary. They have experience with establishing  meaningful educational experience for their students.  Many faculty advisers work with local engineering firms and various government agencies. This is a tremendous resource. Department advisory committee members working with faculty members will no doubt be excited about this endeavor. They may find working with faculty and students by sharing information with GitHub and other open-source server providers beneficial to themselves, their customers and   students.


__Package Summary__ 

A quick read of the __Home Analytical Tools Section__ may have given the impression that the goal to create the __cartools__ package, a collection of __R__ computer algorithms, will be overwhelming difficult to understand and implement. Words, like complex, dynamics, theory, parsimonious, stochastic and model integration can be daunting. No question that designing and operating a transportation system are difficult problems. Hopefully, by dissecting the problem into manageable pieces and synthesizing the  results from the individual pieces brings clarity to the process. The ultimate aim of using __cartools__ package tools is to improve our chances of mitigating congestion and improving performance using the simple mathematical and graphical tools. 


__Lessons Learned__

_What have we learned about traffic performance using the __cartools__ package?_ The most notable lesson learned is:

* By modeling driver behavior with a simple, parsimonious  Brownian Bridge model of speed, it is possible to explain and forecast  traffic  breakdown on a ring-road and at a bottleneck.  

The potential of the __cartools__ package as a learning tool has been demonstrated. Hopefully, this platform can be adapted and expanded to help users  identify the root cause or causes for other congestion problems. 

My next task is add HTML widgets and animation  to the package. Adding these features will allow the user to simply conduct sensitivity analyses on-line. For example, the effects of speed  and density on ring-road performance can be explored. Sliders for $u$, $\sigma$ and $k$ will be provided so a user change them at will. The __cartools__ package will instantaneously produce a histogram of head-ways as illustrated above.

This package is open to the public, https://github.com/PJOssenbruggen/Basic. In other words, I am inviting people to share their experiences with the package with me. See https://guides.github.com/activities/forking/. 

GitHub promotes this activity. See https://github.com/explore. 

__Reference__
