---
title: "Smart Mobility"
bibliography: CF.bibtex
---

***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE, warning = FALSE, message = FALSE, error = FALSE}
library(Basic)
```




As we carefully review the outcomes of the simulations derived from various deterministic and stochastic models, we must conclude that self-optimizing drivers are working against their own self interest in avoiding congestion and minimizing delay. Contrast the results from the __Zipper Merge__ and __Traffic Breakdown Part 7__ simulations. It is instructive. 

The __Zipper Merge__ simulation assumes that drivers act in concert to align themselves before merging. The analysis suggests that drivers can successful merge at any speed, suggesting that high speed merges can take place.  There is no empirical evidence to support this contention other than observing drivers on a  race track.

__Traffic Breakdown Part 7__ simulations assume that drivers act individually. Drivers have no incentive to give their counterparts room to maneuver. As a result, they promote delay. This kind of self-serving behavior is not new to traffic operations. It is known as _Braess's Paradox_ [@braess]. It has been demonstrated both with mathematical models and with empirical evidence that closing certain roads in a roadway network can improve service.    @Roughgarden:2002 reported that total delay using convex programming for a simple idealized "optimum" network would decrease delay by one-third over a "non-optimum" network. 

Critics may claim that counter intuitive methods as suggested by the _Braess's Paradox_ approach has  no application to traffic breakdown at a bottleneck.  It is a far-fetched notion to assume the geometric layout of an interstate freeway interchange, say,  can be improved in any meaningful or practical way. If so, it will be  extraordinarily expensive.  

Secondly, training drivers to change their habits is social engineering  requiring unobtainable behavioral changes. For example, drivers are taught that the right and left freeway lanes are reserved for slower and higher speed travel, respectively. Take any freeway of your choosing and  observe that this teaching is not enforced by law enforcement and  ignored by the majority of the driving public. Tailgating is common practice. 

Finally, critics can claim that the deterministic  __Zipper Merge__ model disregards traffic volatility, therefore it ignores a most fundamental premise of the work presented here in the __cartools__ package. These are formidable criticisms. 

__Intelligent Transportion Systems__ 

Consider a proof-of-concept project. Control traffic noise while leaving  individual driver  behavior more-or-less alone. 

_How can this be done?_ Our answer is  to apply  _Intelligent Transportion Systems_ (_ITS_) in a _smart city environment_. The project consists of three major components:

1. __Collect real-time data.__ Collect data  with lidar and radar equipment placed roadside or collected on-board a vehicle with a GPS system. Measures, including location $x_{vehicle}(t)$ and speed $u_{vehicle}(t)$  data will be transmitted using wireless technology and stored on a central controller computer.

2. __Develop a  real-time forecast algorithm.__ The algorithm will be created along the lines of the __Zipper Merge__ and __brktrials2__ models. In lieu of simulating zipper and traffic breakdown merges, a __brkzipper__ algorithm would be used to provide real-time information  directed at the _individual driver_. "Optimum derived" ${x}_{driver}(t)$ locations and  ${u}_{driver}(t)$ speeds that minimize delay for all drivers. For example, instruct a driver, who will triggered breakdown in the  __Traffic Breakdown Part 7__ simulation, say, to maintain his or her speed and not decelerate. Other drivers would be given similar instructions. By providing individualized instruction, an optimum solution is obtained.

3. __Use  GPS to transmit individualized "optimized" driver instruction.__ To be effective, drivers will relinquish control of their vehicles. GPS coupled with a on-board vehicle control system will be used. The on-board system is contemplated to be similar to  an autonomous cruise control (ACC) device, a device found on contemporary vehicles sold today. This is not social engineering. It is common sense. The driving public applauds, purchases and uses this kind of  innovative technology.

__Smart Cities and the Internet of Things__

"A smart city is an urban area that uses different types of electronic data collection sensors to supply information which is used to manage assets and resources efficiently. https://en.wikipedia.org/wiki/Smart_city" Our proof-of-concept meets this definition. 

However, a smart city involves more than collecting, analyzing and distributing data. While proof-of-concept project involves working with data for sure, @albino state that there is  __a  clear need to engage policy makers from local communities.__ Take the _Uber Driverless-Car-Project_ in the City of Pittsburgh for example https://www.nytimes.com/2017/05/21/technology/pittsburgh-ubers-driverless-car-experiment.html. An Advanced Technology Center was opened for driverless car research, a benefit. However,  Uber failed to honor its promises to form a meaningful partnership with the city.  Uber promised to hire Hazelwood residents, provide seed money and help apply for a federal grant, share traffic data collected by lidar devices mounted on their cars, among other things. No Hazelwood residents were hired, the grant was denied, and the data delivered to the city was insufficient.

__Student Learning:__ We feel that our proof-of-concept project provides, at minimum, a good learning experience for students in engineering and computer science http://eng-cs.syr.edu/transformation/. It can be extended to be an excellent multidisciplinary effort where students from various disciplines work together. Two article that appear in the _Syracuse University Spring Magazine_ 2018 issue describe exciting learning opportunities open to students provided by the _Maxwell X Lab_, _iSchool_ and _Whitman School_ https://www.syracuse.edu/wp-content/uploads/sumagazine-2018-spring.pdf. 

__Data Sharing:__ In this age of _Big Data_, it is appropriate for public, private and educational institutions to share their knowledge and data resources. If nothing else is gained from the  __cartools__  website, a reader must realize good field data are essential. It may seem surprising that the data used in our studies is special. Why? It was continuously measured _24/7_ over an eight month period. As importantly, it contains speed data. Typically, DOTs around nation do not collect speed data and only  flow data [@nydot]. There are no speed data in _CNY_.  

__Cooperative Learning:__ We have established a dialog with University of Tennessee in Chattanooga, TN. "The Underground Infrastructure Sensing project, headed by UTC’s Dr. Dalei Wu in the College of Engineering and Computer Science, was chosen by Smart Cities Connect Media and Research for the award. The project is a collaboration with Dr. Dryer Huston at the University of Vermont." https://blog.utc.edu/news/2018/03/utc-named-one-top-50-green-projects-country/. _UTC_ has recently embarked on a data collection effort on a major thoroughfare in the city as described above. See __Collect real-time data__. 

One of these days, I-81 will be replaced. Traffic data could be used in a before-after study to determine the benefits obtained from replacing it. It could also prove to be of great value for planning to use _smart mobility systems_ in Syracuse, undeserveably  whose national reputation is marred by being located in the U.S. Rust Belt. 

__The Academic Mission:__ Since no one university has all the resources at this point in time (maybe never) to tackle a _Big Data_ project, we are reaching out to form  partnerships with _SU_, _UTC_ and _UVM_. As a starter, we would like discuss the possibility of talking with professors who will be working on senior projects for the spring semester of 2019. We know a few faculty advisers who work with local engineering firms and various government agencies. This is a tremendous resource.  They will find it rewarding and may also discover that sharing information with GitHub and/or other open-source servers gives   them and their    students  a valuable tool.

Denise Owen Harrigan states in her _Syracuse University Spring Magazine_ article that "At Syracuse, Big Data is a big player. It’s steering academic programs in new directions, igniting interest throughout campus, and inspiring new research questions and innovative research methodologies." We believe that our proof-of-concept project meets this criteria.


*** 
__References__
