---
title: "Highway Performance"
bibliography: CF.bibtex
---

__Model Types__

We have developed two modeling types to simulate traffic behavior.  Each type realisticly simulates conditions found in the field  using $t-x$ trajectories as the principle means of comparison. The models  are fundamentally different. One is a deterministic model and the other is a stochastic model. The forecasts from the two models compliment one another. Both satisfy the principal aim of developing the __cartools__ package: to explain highway performance as simply as possible.  

```{r, include = FALSE, warning = FALSE, message = FALSE, error = FALSE}
library(Basic)
```



``` {r table 1, echo = FALSE, message=FALSE, error=FALSE, warning = FALSE}

df1  <- data.frame(c("Microscale time-series", "No", "Sight lines","Safe headway", "No"))
df2  <- data.frame(c("Microscale time-series", "Yes", "Noise","Safe headway", "Yes"))
df   <- cbind(df1,df2)
colnames(df) <- c("Acceleration Model", "Brownian Bridge Model")
rownames(df)  <- c("Description", "Stochastic", "Driver Response", "Driver Risk Aversion", "Empirical Data Link")
knitr::kable(df,  align = "c")
```

The major strength of the  __acceleration model__ is its ability to explain how a driver will vary his or her vehicle speed over time owing to traffic conditions. It features a $\ddot{x}(t)$ model, a _non-linear, deterministic model_. 

The major strength of the __Brownian bridge model of speed__, a stochastic model $U(k,t)$, is its ability to explain the role of traffic noise and role that traffic density $k$ plays in forecasting the probability of being in a congested state, $\pi(k).$ Since the $U(k,t)$ model is derived from a controlled experiment where $k$ is a fixed value, traffic flow, a most important measure of highway performance, is simply defined as a _parsimonius, stochastic model_:

* $Q(t)  = k * U(k, u, \sigma, h_{safe}, t)$.

To draw attention to the significance of the model in explaining traffic flow with $k$, we rewrite as:

* $Q(t)  = k * U(t)$

where $U(t) = U(k, u, \sigma, h_{safe}, t)$.



__Brownian Bridge and Bottleneck Problem__ 

_I am working of a Brownian bridge computer algorithm that explains traffic breakdown (the freeway lane drop problem)._




__Package Summary__ 

A quick read of the __Home Analytical Tools Section__ may have given the impression that the goal to create the __cartools__ package, a collection of __R__ computer algorithms, will be overwhelming difficult to understand and implement. Words, like complex, dynamics, theory, parsimonious, stochastic and model integration can be daunting. No question that designing and operating a transportation system are difficult problems. Hopefully, by disecting the problem into manageable pieces and synthesizing the  results from the individual pieces brings clarity to the process. The ultimate aim of using __cartools__ package tools is to improve our chances of mitigating congestion and improving performance using the simple mathematical and graphical tools.


__Lessons Learned__

_What have we learned about traffic performance using the __cartools__ package?_ The most notable lesson learned is:

* By modeling driver behavior with a simple, parsimonious  Brownian Bridge model of speed, it is possible to explain and forecast  traffic  breakdown on a ring-road and at a bottleneck.  

The potential of the __cartools__ package as a learning tool has been demonstrated. Hopefully, this platform can be adapted and expanded to help users  identify the root cause or causes for other congestion problems. 

My next task is add HTML widgets and animation  to the package. Adding these features will allow the user to simply conduct sensitivity analyses on-line. For example, the effects of speed  and density on ring-road performance can be explored. Sliders for $u$, $\sigma$ and $k$ will be provided so a user change them at will. The __cartools__ package will instantaneously produce a histogram of headways as illustrated above.

This package is open to the public, https://github.com/PJOssenbruggen/Basic. In other words, I am inviting people to share their experiences with the package with me. See https://guides.github.com/activities/forking/. 

GitHub promotes this activity. See https://github.com/explore. 

