---
title: "Highway Performance"
bibliography: "CF.bibtex"
---

__Model Comparison__

We have developed two modeling types to simulate traffic behavior.  Each type realisticly simulates conditions found in the field, i.e., using $t-x$ trajectories as the principle means of comparison. Both approaches explain traffic behavior. The approaches compliment one another using fundamentally different mathematical models. The following  table gives a brief summary of two approaches. 

```{r, include = FALSE, warning = FALSE, message = FALSE, error = FALSE}
library(Basic)
```



``` {r table 1, echo = FALSE, message=FALSE, error=FALSE, warning = FALSE}

df1  <- data.frame(c("Microscale time-series", "No", "Sight lines","Safe headway", "No"))
df2  <- data.frame(c("Microscale time-series", "Yes", "Noise","Safe headway", "Yes"))
df   <- cbind(df1,df2)
colnames(df) <- c("Acceleration Model", "Brownian Bridge Model")
rownames(df)  <- c("Type", "Stochastic", "Driver Response", "Driver Risk Aversion", "Empirical Data Link")
knitr::kable(df,  align = "c")
```

The major strength of the  __acceleration model__ is its ability to explain how a driver will vary his or her vehicle speed over time owing to traffic conditions. It features a non-linear $\ddot{x}(t)$ model. 

The major strength of the __Brownian bridge model of speed__, a stochastic model $U(k,t)$, is its ability to explain the role of traffic noise and role that traffic density $k$ plays in forecasting the probability of being in a congested state, $\pi(k).$ Since the $U(k,t)$ model is derived from controlled experiment where $k$ is a fixed value, traffic flow, a most important measure of highway performance, is simply defined as a _parsimonius, stochastic model_:

* $Q(t)  = k * U(k, u, \sigma, h_{safe}, t)$.

To draw attention to the significance of the model in explaining traffic flow with $k$, we rewrite as:

* $Q(t)  = k * U(t)$

where $U(t) = U(k, u, \sigma, h_{safe}, t)$.



__Brownian Bridge and Bottleneck Problem__ 

_I am working of a Brownian bridge computer algorithm that explains traffic breakdown (the freeway lane drop problem)._




__Package Summary__ 

A quick read of the __Overview__ may give the impression that the approach, a package of computer algorithms, will be difficult to understand and overwhelming to implement. Words, like complex, dynamics, theory, parsimonious, stochastic and model integration can be daunting. No question that designing and operating a transportation system are difficult problems. By disecting the problem into manageable pieces and synthesizing the  results from the individual pieces brings clarity to the process. 


The mathematics behind the animation is relatively simple. Of course, the Brownian motion model shown here can be adapted to other situtations. This will be  illustrated presently. The ultimate aim of using [trfhp](https://github.com/PJOssenbruggen/Basic) package tools is to improve our chances of mitigating congestion and improving performance using the simple mathematical and graphical tools.


__Lessons Learned__

_What have we learned about traffic performance using the __cartools__ package?_ The most notable lesson learned is:

* By modeling driver behavior with a simple, parsimonious  Brownian motion model of speed, it is possible to explain and forecast  traffic  breakdown on a ring-road and at a bottleneck.  

The potential of the __cartools__ package as a learning tool has been demonstrated. Hopefully, this platform can be adapted and expanded to help users  identify the root cause or causes for other congestion problems. 

My next task is add __R Markdown Shiny__ HTML widgets to my code. Adding this feature will allow the user to simply conduct sensitivity analyses on-line. For example, the effects of speed  and density on ring-road performance can be explored. Sliders for $u$, $\sigma$ and $k$ will be provided so a user change them at will. The __cartools__ package will instantaneously produce a histogram of headways as illustrated above.

This package is open to the public, https://github.com/PJOssenbruggen/Basic. In other words, I am inviting people to share their experiences with the package with me. GitHub promotes this activity, https://github.com/explore.

