---
title: "Explaining Traffic Breakdown"
bibliography: CF.bibtex
---

***
[^1]: The __cartools__ functions __usdzipper__, __formqueue__, __freeflowpass__ __passplot__ __xabmerge3__, __brktrials3__ and __brkdelay__ are used to create the plots shown on this page.  


__Model Types__

We have developed two modeling types to simulate traffic behavior.  Each type realistically simulates conditions found in the field  using $t-x$ trajectories as the principle means of comparison. The models  are fundamentally different. One is a deterministic model and the other is a stochastic model. The forecasts from the two models compliment one another. Both satisfy the principal aim of developing the __cartools__ package: To explain highway performance as simply as possible, two model structures have been used:  

```{r, include = FALSE, warning = FALSE, message = FALSE, error = FALSE}
library(Basic)
```



``` {r table 1, echo = FALSE, message=FALSE, error=FALSE, warning = FALSE}

df1  <- data.frame(c("Microscale time-series", "No", "Sight lines","Safe headway", "No"))
df2  <- data.frame(c("Microscale time-series", "Yes", "Noise","Safe headway", "Yes"))
df   <- cbind(df1,df2)
colnames(df) <- c("Acceleration Model", "Brownian Bridge Model")
rownames(df)  <- c("Description", "Stochastic", "Driver Response", "Driver Risk Aversion", "Empirical Data Link")
knitr::kable(df,  align = "c")
```

A major strength of the  __acceleration model__ is its ability to explain how a driver will vary his or her vehicle speed over time owing to traffic conditions [@LiliBook]. It features a $\ddot{x}(t)$ model, a _non-linear, deterministic model_. A major strength of the __Brownian Bridge model__ is its ability to capture traffic noise. 

By connecting these two models, we can explain traffic breakdown and simulate traffic breakdown events. To appreciate the challenges of  developing our stochastic breakdown model, we follow a step-by-step approach and at the same time, we highlight the special features of each model type. Now, we want synthesize this information in a coherent manner. The final model or __brktrials3__ function, incidentally, is described and critiqued in  __Traffic Breakdown Part 7__. 

_What is a traffic breakdown event?_  It is as the transition from a free-flow state to a congested state, $X = 1 \rightarrow X = 0$. The state space notation is simple enough and should be clear. The groundwork for explaining a traffic breakdown event with the two model structures was established earlier. See menu items: __Drivers__, __Noise__, __Ring-Road__ and __Bottleneck__. 



__Traffic Breakdown Part 1__ 


Explaining traffic breakdown with a stochastic  model is most challenging. Why? A traffic breakdown event is complicated because there are a myriad of ways a breakdown event can be initiated. Weaving, merging, aggressive driving, and poor roadway design are often blamed. To simplify the discussion, we assume that traffic breakdown will occur at a bottleneck. See the schematic diagram in __Bottleneck__. A  two-lane freeway with uni-directional flow merges into a single lane, the so-called "lane drop" problem. Vehicles are forced to merge. For the present, we concentrate on the first two vehicles that decelerate in traffic described to be highly volatile with high density. 

There is uncertainty associated  specifying a  stochastic speed model. Since the __acceleration__ or deterministic $u(t)$ model explains breakdown, we will add random effects component to it [@iacus]. We obtain:

* $U(k, a, b, h_{safe}, t) =  u(0) +  a * t - \frac{b}{2} * t^2 + W(\Delta t)$

where $W(\Delta t) \sim N(0, \sigma \Delta t)$ and  $\sigma = \sigma(k)$. Thus, $U(k, a, b, h_{safe}, t)$ is now a stochastic model with an external data link.  Does it explain breakdown?

Here is the scenario. The following $t-u$ and $t-x$ plots are derived from  a simulation using the $U(k, a, b, h_{safe}, t)$ model.[^1] The simulation is limited to an investigation  of two vehicles, a lead vehicle and a following vehicle, identified as  1 and 2. It is assumed that the drivers of these vehicles will not change lanes. They car-follow exclusively. There are no drivers outside their driving lane. Thus, conditions that describe a zipper merge do not apply here.  Even with these simplifying assumptions, there is quite a bit going on during a traffic breakdown. 

```{r usdzipper,  message = FALSE, fig.height = 8, fig.width = 10, echo = FALSE}
par(mfrow = c(1,1))
set.seed(124)
output1 <- usdzipper(0, 5, 41, 11.6, 0, 500)
```

Since  traffic is highly volatile, the first step is to simulate field conditions. We assume the traffic density is fixed, $K = k$ = 50 vpm. Thus $\bar{u} \pm \sigma = 41 \pm 11.6$ mph. Since $X = 1 \rightarrow X = 0$, the speeds of the two vehicles before and after breakdown must satisfy the constraints $u(t_{start}) > u^*$ and $u(t_{end}) \le u^*$ where the time at the start and end of the simulation are $t_{start}$ = 0 and $t_{end}$ = 5 seconds. These values were estimated by drawing samples from a simulated data set consisting of one hundred draws from the normal distribution of $N(\bar{u} = 41, \hat{\sigma}^2 = 11.6^2)$. Four draws were made from the simulated data. They are shown in the following table. 

``` {r table 2, echo = FALSE, message=FALSE, error=FALSE, warning = FALSE}

df1  <- data.frame(c("0", "55 mph (81 fps)", "0 feet", "57 mph (83 fps)", "-68 feet"))
df2  <- data.frame(c("5 seconds", "41 mph (60 fps)", "500 feet", "32 mph (47 fps)", "455 feet"))
df   <- cbind(df1,df2)
colnames(df) <- c("Start", "End")
rownames(df)  <- c("Time t", "Vehicle 1:  u", "Vehicle 1:  x",  "Vehicle 2:  u", "Vehicle 2:  x")
knitr::kable(df,  align = "c")
```

Note that at the start of the simulation, $t_{start} = 0$, that vehicle 2 is traveling at a speed slightly greater than vehicle 1. At the end of the simulation, $t_{end}$ = 5 seconds, the opposite is true.  The simulation also shows driver 2 is traveling at higher speed than driver 1 for the first three seconds. Driver 2 begins to  decelerate around three seconds to maintain a safe headway at $t_{end}$. Speed volatility explains these findings. Another simulation will be different.

The second step is to estimate model parameters $a$ and $b$ for each of the two vehicles. Other assumptions must be must made. The bottleneck is assumed tapper down from two to one lane in 500 feet and the total time for the breakdown is 5 seconds. The drivers are assumed to be risk averse so driver of vehicle 2 maintains a safe headway. All this information is summarized in the table. The $t-x$ trajectory shows vehicle 2 maintains a safe headway for the entire 5 seconds of the simulation. 

The same numerical integration scheme described earlier for determining $x_t$ trajectories for the ring-road is used here. Like the ring-road, the $t-x$ trajectories are smooth, suggesting the speeds at times $t_0$ and $t_{end}$ are more critical than the speed changes caused by Brownian motion $W(t)$ ranging over the interval $t_{start} < t < t_{end}$. 

__Traffic Breakdown Part 2__ 

The importance of maintaining a safe headway cannot be overemphasized. Without this assumption, the $t-x$ trajectory of vehicle 2 could cross the trajectory for vehicle 1. The result is a rear-end crash caused by _tailgating._ This possibility of  a rear-end crash can occur even when there is a safe headway at $t_{start}$ = 0 and $t_{end}$ = 5. Let's see.

The following simulation shows a near-miss caused by _tailgating._ While the trajectories do not cross, the $t-x$ trajectories shows a safe headway violation between three and four seconds. Since drivers are assumed to risk averse, it is not unreasonable to assume the driver of the following vehicle will decelerate. In other words, not violate the safe headway assumption. To keep the discussion as simple as possible, the safe sight-line assumption for times $0 < t < t_{end}$, discussed in the __Bottleneck__ section,  is not imposed here. 


```{r usdzipper2,  message = FALSE, fig.height = 5, fig.width = 10, echo = FALSE}
par(mfrow = c(1,1))
set.seed(129)
output1 <- usdzipper(0, 5, 41, 11.6, 0, 500)

```

__Traffic Breakdown Part 3__ 

How is a traffic breakdown initiated in the first place? So far, we have tacitly assumed the lead vehicle decelerates from $u_{start} > u^*$ and $u_{end} < u^*$. Now we tackle a most challenging question. Instead of assuming to watch vehicles pass us from the road side, we will position ourselves overhead by analyzing photographs taken from a helicopter or drone flying overhead. 

```{r formqueue,  message = FALSE, fig.height = 5, fig.width = 8, echo = FALSE}
par(mfrow = c(1,2))
set.seed(129)
output1 <- formqueue(0, 10, 55, 11.6, -1000, -500, 14)
```

Here, the following graph shows two lanes of traffic merging into one lane. There is no  traffic breakdown in this simulation. No queue forms. All drivers maintain free-flow speeds,  $u > u^*$ where $u = \bar{u}$ = 55 mph. As above, we simulated speed using the $U(k, a, b, h_{safe}, t)$ model. To maintain a $X = 0$ state, all drivers of following vehicles keep a safe headway. This result is most vividly seen downstream where $x > 0$ where the vehicles travel in one lane. 

Since is controversy associated with the proper positioning of roadside equipment, it is interesting to note the following:

* $Q = Q(x > 0) = Q(x \le 0)$

because $k$ = 50 vpm is a constant and $\bar{u}$ is the same value both downstream and upstream  of $x = 0$. Thus, $Q = k * \bar{u}$ where $k = \frac{n}{l}$ and $n$ is the total vehicle count obtained from the photo. A roadside collection of data taken any position on the photo downstream or upstream of $x = 0$ gives the same result. We look at the controversy again after we answer the question: How is a traffic breakdown initiated in the first place?

__Traffic Breakdown Part 4__ 

The __Zipper Merge__ graph  demonstrates the importance of upstream vehicle alignment  and the effects of noise of  merging vehicle at a bottleneck over time. The spacing between  vehicles is nonuniform and vehicle grouping is virtually impossible to predict. 

A zipper merge is a special case because the upstream vehicles are aligned in a manner that it is unnecessary for vehicles to change speed. Now, we eliminate this assumption and put drivers under the stress by requiring a speed change of only one vehicle. Since our analysis takes place where the chance of a traffic breakdown is significant, $k = 50$ vpm, the action of this driver may trigger a breakdown event causing a queue to form that moves upstream from $x$ = 0.    

To simplify the discussion, the effect of noise will be temporarily removed from the following  "thought experiment." This step may seem odd given the emphasis placed on stochastic modeling. After we complete the analysis, we will be in a better position to understand breakdown. More specifically, explaining and coming closer to reliably predicting  the role that traffic noise plays in triggering breakdown. In other words, determine if traffic noise is a major or minor factor in explaining and predicting breakdown.

We will continue to observe vehicles from overhead. But this time, we will follow the actions of only two drivers, who are traveling side-by-side in lanes 1 and 2 at  $x = -500$   feet, the beginning of the merge section where freeway tappers down from two to one lane. The side-by-side vehicles, like the rest of the vehicles traveling in the direction of the bottleneck, are traveling at speed $u$ = 55 mph. The driving scenario is similar to one described above before being transformed from a free-flow to congested state, $X = 0 \rightarrow X = 1$. 

Since the two drivers are side-by-side, one must make the decision to be the leader or follower. The choices are: (1)  __accelerate and merge__ and become a leader, or (2) __decelerate and merge__ and become a follower. In either case, the drivers will do so safely. They will  maintain a safe headway of at least $h_{safe} = 53$ feet for speed of $u$ = 41 mph. 
Most importantly and regardless of choice, the decision-maker does not want to disrupt the speeds of the other drivers, which are assumed to be traveling at $u$ = 41 mph. The aim is to cause no further disruption or delay. In essence, the decision-maker wants to perform a zipper merge.


A series of photographs are taken  for the two drivers in the two cases.  Speed and location recordings are made  every second over a period of 50 seconds. The data are analyzed and summarized as shown below in the $y-x$ plots. Since the location $x$ is measured in feet, speed $u$ is shown in feet per second (fps).


```{r freeflowpass,  message = FALSE, fig.height = 4.5, fig.width = 4, echo = FALSE}
par(mfrow = c(1,2))
library(glm2)
library(ggplot2)
library(Basic)
df1 <- freeflowpass(0, 10, 55, 0, -1000, -500, 14)
passplot(df1, title = "Accelerate and Merge")


df2 <- decelmerge(0, 10, 55, 0, -1000, -500, 14)
passplot(df2, title = "Decelerate and Merge")


```

Regardless of the action taken, the aim of the decision-maker is achieved. The driver does not disturb the other drivers. No queue is formed. The __accelerate and merge__ driver reaches a maximum speed of 50 mph (103 fps) and the __decelerate and merge__ driver reaches a minimum speed of 40 mph (58 fps). The acceleration and deceleration rates are approximately equal to 7  and -7 ft/sec$^2$, respectively. A maximum comfortable decelerate is considered to be 15 ft/sec$^2$. Since all drivers are maintaining safe head-ways,  no crash or near miss incidents are predicted. 


__Traffic Breakdown Part 5__ 

Ignoring traffic noise is an ideal scenario and totally unrealistic. There is little doubt that the root cause traffic breakdown must consider traffic noise. 

There are other factors that must be considered. Take for example, the safe headway $h_{safe}(u, l_{eff})$ assumption. Drivers will not only self optimize but will be risk averse, they seek comfort. We are hard pressed to find a real world situation where two drivers will maintain a safe headway of $h_{safe}$ = 67 feet as shown in the $y-x$ plots $x > 0$, downstream of the bottleneck. Tailgaters may not abide by this rule of keeping three cars lengths between themselves and the vehicle in front of them. Drivers of the lead vehicle will feel more comfortable with $h_{safe}$ = 106 feet, more than six car lengths between the lead and following vehicles, as shown for $x \leq -500$  feet. 

This critique of the deterministic model and subsequent model developments brings into question what are the conditions are necessary for a decision-making driver to take action. For $-500 \le x \le 0$ feet, it was assumed that the safe headway $h_{safe}(u, l_{eff})$ assumption was always sufficient. A decision-making driver, who is not comfortable with the rule, has the option to stop and trigger a breakdown event. The challenge is to develop a stochastic model of the form:

* $U(k, a, b, h_{safe}, h_{comfortable}, t)$.

__Traffic Breakdown Part 6__ 

To trigger a traffic breakdown and to initiate queue formation, a roadway system must be under stress. Even though we are investigating a simple roadway geometry, the system is under stress: (1) the traffic density is high, $k$ = 50 vpm, (2) vehicles are forced to merge because of a lane drop, (2) traffic volatility is high, $\sigma$ = 11 mph = 16 fps, for an average speed $u$ = 41 mph = 60 fps, (3) vehicles are closely spaced thus limiting  drivers' options to maneuver and finally, (4) drivers desire to self optimize work against them putting the system under added stress. In the following simulation, two  of  three vehicles are traveling side-by-side before they enter the merge zone of a bottleneck. To avoid a crash, one driver must yield and decelerate. Given this combination of factors, traffic breakdown is expected. The chance of a breakdown is $\pi(k = 50)$ = 0.8. Once breakdown is triggered, a queue forms and grows steadily over time.    
 
Let us look at the action of the three drivers, who we identify by vehicle number. Their individual action  over time is important. Initially at time $t_0$ = 0 seconds, the distance between vehicles 0 and 1, which is directly behind, is $h_0$ = 105 feet, a space headway estimated as $h_0 = 1/k$. $h_0$ is almost twice the safe headway $h_{safe}$ for an average speed of $u$ = 41 mph = 60 fps at this density. The positions of the three vehicles at $t_0$ = 0 seconds is shown under the table column heading $x_0$. The table shows there is no conflict between vehicles 0 and 1 and indicates that vehicle 0 can move freely downstream through the merge zone, $-500 < x \le 0$. Since the driver has no conflicts, we call it a _zipper merge_. The slope of the $t-x$ trajectory appears to be steady. However, the table shows the driver to be increasing his or her speed. At times $t_{start}$ = 6 seconds and $t_{end}$ = 14.5 seconds, the vehicle has accelerated from $u_{start}$ = 64.7 fps and $u_{end}$ = 74.4 fps.  Times $t_{start}$ and $t_{end}$ are the times when the vehicle enters and exits the merge zone. 


``` {r table 3, echo = FALSE, message=FALSE, error=FALSE, warning = FALSE}
df7  <- data.frame(c("0", "1", "2", "2"))
df9  <- data.frame(c(-895, -1000, -1000, -1000))
df0  <- data.frame(c("No","Yes","Yes","Yield"))
df8  <- data.frame(c("_Zipper_","_Leader_","","_Follower_"))
df1  <- data.frame(c("6", "8.5", "8", "8"))
df2  <- data.frame(c("14.5", "16", "16.5", "16"))
df3  <- data.frame(c("64.7", "62.2", "81.9", "81.9"))
df4  <- data.frame(c("74.4", "53.0", "56.9", "53.0"))
df5  <- data.frame(c("-515", "-501", "-508", "-508"))
df6  <- data.frame(c("-27.3", "-26.4", "-28.9", "-91.0"))
df16   <- cbind(df7,df9,df0,df8,df1,df2,df3,df4,df5,df6)
colnames(df16) <- c("Vehicle","$x_{0}$", "Conflict","Resolution", "$t_{start}$", "$t_{end}$","$u_{start}$", "$u_{end}$","$x_{start}$", "$x_{end}$")
knitr::kable(df16,  align = "c")
```

The situation for drivers of vehicle 1 and 2 is different. There is no conflict upstream of the bottleneck for $-1000 < x < -500$ feet. The trajectories over this range overlap but there is no conflict because the vehicles are traveling in separate lanes. After $x$ = -500 feet, a conflict exists. Vehicle 1 becomes the _leader_ of this pair of vehicles. Its trajectory is similar to vehicle 0 in the sense that the driver travels forward unimpeded. Note its speeds through the merge zone are $u_{start}$ = 62.2 fps and $u_{end}$ = 53.0 fps, respectively.
 

```{r brksummary,  message = FALSE, fig.height = 4.5, fig.width = 6, echo = FALSE}
par(mfrow = c(1,1))
library(glm2)
library(ggplot2)
library(Basic)
set.seed(123)
lst <- brksummary(50, 30, 41, 11, -1000, -500,  14, 0, 0.5)
title(main = "Breakdown")

```

Vehicle 2, the _follower_, decelerates from a speed of $u_{start}$ = 81.9 fps to $u_{end}$ = 53.0 fps. The latter speed of 53.0 fps is exactly the same speed as vehicle 1 at $t_{end}$ = 16 seconds. The distance between the two vehicles at   this time is the safe headway distance $h_{safe}$ = 64.6 feet. The red and blue solid circles draw attention to this fact. The red $t-x$ trajectory shows the curve to cross the blue trajectory for vehicle 1 at two places. At the first crossing at around $x$ = -250 feet, the following $y-x$ shows the center-line of the two vehicles to be separated by 6 feet. The slopes of the $t-x$ trajectories for vehicles 1 and 2 suggest that vehicle 1 is accelerating and vehicle 2 is decelerating at this point in time.

The red trajectory  shows vehicle 2 to cross the trajectories of vehicles 0 and 1. This does not mean the vehicle is crashing into these vehicles. Its meaning is quite the opposite. It shows that once a vehicle passes the bottleneck $x$ = 0 that drivers tend to accelerate and seek higher speeds. In this instance, the driver of vehicle 2 is passing vehicles 0 and 1 in the opposing lane that is free of traffic.

```{r passplot,  message = FALSE, fig.height = 4.5, fig.width = 5, echo = FALSE}
par(mfrow = c(1,1))
library(glm2)
library(ggplot2)
library(Basic)
set.seed(123)
#passplot(df, title = "Breakdown")
#print(lst[[2]])
```

To keep the discussion as simple as possible, we investigated three vehicles. Since their trajectories satisfy the safe headway conditions, all three drivers are driving safely. Now, imagine that vehicle 3 and 4 enter the simulation. Given that the two self-optimizing drivers will maintain a safe head-ways between themselves and the vehicles in front them as we described for car-following. These drivers also have the option to maintain a $1/k$ headway described above. In either case, vehicles 3 and 4 are suspected to driving side-by-side especially as they approach $x = -500$ feet. Using the $1/k$ headway estimation method, they will be located at $x = -1105$ feet at $t = 0$ and the two drivers of vehicles 3 and 4 will be in a position to decide which one take on the role of leader or follower. In other words, we can repeat the analyses as above for vehicles 1 and 2. The net result is the system will remained in a stress condition. The queue will grow and speeds will continue to decline in magnitude. Eventually, the speeds will decline to values to values that we define as congested. We experience the transition, $X = 0 \rightarrow X = 1$.

__Traffic Breakdown 7__

Now, we will introduce another vehicle to our simulation.  Why restrict it to four vehicles? Adding more vehicles is no problem but the $t-x$ trajectories for more than four vehicles becomes different to follow owing to the complexity associated with traffic breakdown. __cartools__ is an educational and research tool. Right now, the emphasis is on education so we want a problem that is easy to follow  and at the same time, give a clear-cut explanation of traffic breakdown. 

__Model Features and Warnings:__ As we have seen, $t-x$ trajectories  overlap and cross one another. This is permissible in certain situations. We impose car-following restrictions on all vehicles (1) upstream of $x$ = -500 feet where the roadway has two traffic lanes and (2) downstream of $x$ = 0 where the roadway has one lane.  (3) While no car-following restrictions are imposed in the merge zone, $t-x$ trajectories are expected to diverge when the $h_{safe}$ rule is violated.

A reader should be aware that a $t-x$ trajectory is an imperfect tool and it can be misleading when  trajectories  overlap. Look at the first two items. They may seem to have counter intuitive explanations. But really, they are not:

(1) _Upstream $t-x$ trajectories may overlap._ Why? The vehicles are traveling side-by-side and do not occupy the same space.

(2) _Downstream $t-x$ trajectories may not overlap._ Why? Overlapping trajectories suggest that  vehicles have crashed. Moreover, we are looking for downstream trajectories are sufficiently far enough apart to assure that the drivers are driving safely. 

While these restrictions seem harmless enough, they place an extra burden on the development of stochastic models. Fortunately, we  analyze  interaction  leader and follower pairs, thus the analysis becomes more palliative. 

__"Desire-Line" Trajectories:__ At the same time restrictions are placed on traffic operation, we want to allow each driver to be a self optimizer and to seek a path to minimize his or her travel time.  To come to grips with this idea, we introduce the notion of a driver "desire-line" trajectory, an ideal scenario for a driver traveling in free-flow or $X = 0$ state. Accordingly, the $t-x$ trajectory the driver follow as predicted by the __bmfree2__ function. This idea will become clearer as we proceed.

__Problem Statement:__  Assume that the traffic entering the bottleneck is operating in a free-flow $X = 0$ state with traffic densities on lanes 1 and 2 equal to $k_1$ = $k_2$ = 25 vpm. Owing to the lane drop, the downstream density will be double, $k_{downstream}$ = 50 vpm. According to a previous discussion,   an average downstream speed of $u_{downstream}$ = 41 mph is expected. Since $u_{downstream}$ < $u^*$ = 50 mph, the definition of a congested state $X = 1$ is satisfied. We will see if this will happens.

However, remember that being in a congested state using field data is a chance event. $\pi(k)$ = 0.8 = 80$\%$ at $k$ = 50 vpm. Stated another way, downstream vehicles are expected to operate at speeds equal to and greater than 50 mph 20$\%$ of the time. The upstream and downstream $\bar{u} \pm \sigma$ are: 68 $\pm$ 4.4 and 41 $\pm$ 11 mph, respectively.  In other words, we expect a substantial drop in speed and a substantial increase speed volatility measured as a standard deviation when a breakdown occurs.

We also assume that the upstream vehicles are traveling side-by-side in parallel and the condition for a zipper merge is not satisfied. Drivers must adjust their speeds to safely merge. We will investigate this behavior.

_Does our stochastic model adequately explain the triggering of a traffic  breakdown event?_ Our investigations begins by evaluating the  "desire-line" $t-x$ trajectories from one simulation shown below. Assigning a different random seed to the algorithm will give an entirely different set of trajectories. 


```{r bottleneck_merge,  message = FALSE, fig.height = 6, fig.width = 8, echo = FALSE}
par(mfrow = c(1,1))
library(Basic)
library(rlist)
lst      <- brktrials3(4, 68.4, 4.4, 0, 30, -700, 0.25, FALSE, 14, -500)
tuxv     <- lst[[1]]
tuxv.fix <- lst[[2]]
vehorder <- lst[[2]]
xfunnel  <- -500
output   <- plotdesire3(tuxv, xfunnel)
```


__Driver Conflicts:__  The "desire-line" trajectories shown above have vehicles 1 and 2 traveling in lane 1 and vehicles 3 and 4 traveling in lane 2.  As mentioned above, (1) upstream $t-x$ trajectories may overlap, (2) downstream trajectories can neither cross nor overlap, and (3) merge zone trajectories are expected to diverge when the $h_{safe}$ rule is violated. Since the downstream trajectories overlap, it is obvious that the drivers must make adjustments to avoid crashing. However, it is not obvious that the $h_{safe}$ rule is violated in the merge zone. The rule is violated, but why?

Let us investigate driver conflicts by starting with vehicles 1 and 3. The drivers of these vehicles have "desire-line" trajectories that are arguably the same. Both drivers want (desire) to arrive at the same location at time $t$ = 30 seconds. This is impossible because their downstream trajectories overlap. One driver must yield to avoid a crash. It is not obvious at this point which driver will yield. This will be discussed presently.

Now, let us shift our attention to driver 2 of lane 1,  a following vehicle in a car-following situation. Driver 2  is maintaining a safe headway behind  vehicle 1. Similarly, driver 4 of lane 2, a following vehicle, is maintaining a safe headway behind  vehicle 4. 

Like in the previous paragraph, one of the drivers of vehicles 2 and 4 must yield to avoid a downstream crash. The "desire-line" of vehicle 2 shows it driver to be more aggressive than driver 4. Driver 2 wants to pass driver 4. Again, we have a situation where one of these drivers must yield. 

__Model Predictions:__ Now, we skip directly to our solution and then critique the solution and the entire approach.

Vehicle 1 becomes the leader followed by vehicles  3, 4 and 2 in that order at time $t$ = 30 seconds. All driver conflicts are removed and the $h_{safe}$ rule is imposed on all drivers. It is also interesting to observe that driver 4, who was deemed to be more aggressive than driver 2, winds up last. Before investigating this situation, we focus on our main concern, _Does our stochastic model adequately explain the triggering of a traffic  breakdown event?_

```{r bottleneck_merge2,  message = FALSE, fig.height = 6, fig.width = 9, echo = FALSE}
par(mfrow = c(1,1))
set.seed(123)
library(Basic)
library(rlist)
lst      <- brktrials3(4, 68.4, 4.4, 0, 30, -700, 0.25, FALSE, 14, -500)
tuxv     <- lst[[1]]
tuxv.fix <- lst[[2]]
vehorder <- lst[[3]]
xfunnel  <- -500
output2  <- plotmerge3(tuxv.fix, vehorder, xfunnel)
```

By visual inspection, we can see the slopes of the $t-x$ trajectories or speeds to be declining for each vehicle that pass through the merge zone and pass  point $x$ = 0.  Even though we are observing only four vehicles, we can also see a queue forming.  It moves in the upstream direction as expected. So, a cursory test of our model, the __plotmerge3__  function, suggests the model explains breakdown. Since we are using the  _Learning Process_ of  @box, we subject our model to further testing and offer deeper understanding.

__Car Following Effects:__ The effects of imposing the $h_{safe}$ rule and the non-linear acceleration model,

* $\ddot{x}(t)  = a - b * t$, 

for the merge zone, coupled with the random effects of the Brownian motion model, (1) slow vehicles down and (2) affect  the order in which vehicles pass through the merge zone. While the $t-x$ trajectories appear correct, there is a lot going on  in this simple example. _Obviously, the code must be robust to account for various interactions that  drivers must face when merging  in dense and volatile traffic._ 

Regardless on the mathematical rigor, our model must satisfy the test of _common sense._ So let us pose another question: _Do the simulated drivers behave in a manner that is realistic?_ 

Start with drivers 1 and 3. Driver 1 is shown as to be moving through the merge zone unimpeded. Owing the effects of Brownian motion, driver 1 arrives at $x$ = -500 fractions of a second before driver 3. Therefore, driver 3 yields and slows as shown. That makes sense. 

Driver 3 must make quick decisions as a car-follower. Theoretically,  driver 3 observes and processes the roadway information presented to him or her. To help us describe the situation as simply as possible, the driver is assumed to subconsciously  process  the information as summarized in the "desire-lines" trajectories for vehicles 1 and 3. After processing the information, driver 3 knows a crash is inevitable if no changes are made. Driver 3 slows to a speed that equals the speed of vehicle 1 and decelerates in a manner to be a safe distance behind vehicle 1 when it passes through the bottleneck at $x$ = 0. Again, this makes sense. Driver 3 is risk averse.

The decision-making scheme used by driver 3 required by our mathematical model uses six variables of time, speed and location. They include the time the vehicle arrives at $x = x_{start}$ = -500 feet or $t_{start}$ and the time the driver arrives at $x = x_{end}$ = 0 or $t_{end}$. The speed at these times are $u_{start}$ and $u_{end}$.  These variables are used to create the $t-x$ trajectory shown for vehicle 3 with the __xabmerge3__ function.

Drivers 2 and 4 at this point in the discussion are both car-followers of vehicle 3. Both arrive at $x$ = -500 feet safely behind vehicle 3 as determined by the $h_{safe}$ rule. Vehicle 4 arrives at $x$ = -500 feet slightly before vehicle 2, therefore, driver 4 takes over as the leader and driver 2 as the follower. It follows the reasoning above, so it makes sense.

Driver 4 must make a decision based on the roadway information presented as a follower of vehicle 3. Like above, driver 4 will use the "desire-line" trajectory for vehicles 3 and 2 and find an unsafe situation and make adjustments. Driver 4 is predicted to follow as shown above. The next step is to analyze the situation for vehicle 2, a follower, where vehicle 4 is the leader.  Since the same scheme is used by each seceding driver, there is no need to claim that it is passing our common sense test.  

__Downstream Traffic Behavior:__ We stated above that driver of vehicle 4 is aggressive. How can we explain that downstream of the bottleneck at $x$ = 0 that driver 4 is the least aggressive driver?  At this point, our answer relies on car-following effects, vehicle sorting, and vehicle speeds at $x$ = 0. In other words, upstream conditions of the bottleneck at $x$ = 0 dictate. Simply stated, the driver is caught in traffic and blocked. Common sense? Yes.

The following forecasts shown below are derived from the same model and  inputs including the same random seed as above. The only difference is  10 vehicles  are analyzed in lieu of four. In this simulation, the first three vehicles 1, 6 and 2 are not delayed and the following  ones are. The speeds in the merge zone of the followers  appear  to be affected most sharply. Let us see.  


```{r bottleneck_merge3,  message = FALSE, fig.height = 6, fig.width = 9, echo = FALSE}
par(mfrow = c(1,1))
library(Basic)
library(rlist)
lst      <- brktrials3(10, 68.4, 4.4, 0, 30, -700, 0.25, FALSE, 14, -500)
tuxv     <- lst[[1]]
tuxv.fix <- lst[[2]]
vehorder <- lst[[3]]
xfunnel  <- -500
output3  <- plotmerge3(tuxv.fix, vehorder, xfunnel)
```

__Bottleneck Delay:__ We decided to investigate 10 simulations consisting of 10 vehicles for the situation described above. The results are summarized with a time-series plot of speed averages and standard deviations below. 

```{r bottleneck_merge4,  message = FALSE, fig.height = 6, fig.width = 9, echo = FALSE}
par(mfrow = c(1,1))
library(Basic)
library(rlist)
lst      <- brkdelay3(10,10, 0, 30, 68.4, 4.4, -700, -500, 14, 0.25)

```


The average speeds fall below our critical speed of $u^*$ = 50 mph from about seven seconds from 5 to 12 seconds with an estimated maximum speed drop  of 21 mph at $t$ = 7 seconds.  

The speed volatility is substantial over this period. According to predictions, traffic volatility is even higher after 15 seconds, or on the downstream side of the bottleneck. 

For the present, we  defer questions dealing with queue length and congestion relief. These are tough questions that need more thought and field data. 


__Stochastic Model Critique:__ _Do the collection of_ __cartools__ _functions adequately explain this behavior?_ Yes, we think so. Can it be improved? Definitely. _Our intention is to have a fluid work platform. Therefore, corrections and refinements to the_ __cartools__  _are expected and suggestions for improvement are welcome._ 

We understand that further research is needed to better explain downstream traffic behavior. No field data were collected downstream of the bottleneck on I-93, so our driving experiences are used. It is a fact that once drivers pass through a bottleneck that drivers will accelerate and will try to return to a free-flow state. Well, we don't have this situation. Downstream of the bottleneck, our roadway is confined to one lane and does not expand to two lanes.  Regardless, the forecasts from the simulations shown above show this behavior. It passes our common sense test.

For another, the speed drop and high volatility results shown in the __Bottleneck Delay__ plot stem directly mechanisms embedded in  the stochastic model or __brktrials3__ function. We have empirical evidence from I-93  field data that $\bar{u} \pm \sigma$ equals 41 $\pm$ 11 mph. _This information was not introduced into our model or code._   The model forecasts and  empirical evidence match. We are looking into ways to show that this result is not a coindence, but a fact.


__Conclusions__ 

Traffic noise explains delay and queue formation. High traffic density can be blamed for the loss in performance downstream of $x > 0$. 

Remove noise and all drivers benefit as described above by the zipper merge of __Part 3__.   Does the simulation suggest that expanding the number of lanes downstream of the bottleneck to reduce land traffic density is a good solution? Good question. History shows that it is not a good idea. What can be done? Public transportation is one answer.

The simulation suggests that _removing or mitigating traffic noise that the driving  world would be a better place._ We look at that possibility by looking at Intelligent Transportation Systems (ITS) and the notion of Smart Cities.

*** 
__References__