---
title: "Ring-Road Experiment"
bibliography: "CF.bibtex"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include = FALSE, warning = FALSE, message = FALSE, error = FALSE}
library(Basic)
```

__A Controlled Experiment__

We continue our model building effort by simulating conditions observed on a  _ring_ or _loop road_. The [YouTube video](https://www.youtube.com/watch?v=7wm-pZp_mi0) shows $n$ = 22 vehicles [@1367-2630-10-3-033001].  The drivers cannot pass one another or enter or exit the loop.  Each driver was instructed to maintain a same constant speed $u$. The drivers cannot follow instructions. After a few sections, the traffic breaks down, i.e., drivers cannot maintain speed $u.$ 

Why? The video shows the vehicles to be traveling at a reasonable speed. Initially, all vehicles are uniformly distributed around the circle. After a few seconds,  the vehicles begin to interact because the one or more drivers cannot maintain  speed $u$. If we track a vehicle, we can see the speed will vary by a slight amount. Since their no room to move laterally, the drivers must follow and react to the situation directly in front of them. This behavior is a classicly called  _car-following._ 

It is hypothesized that the vehicles described in the preceding paragragh are operating  in a _congested state_. The spacing between vehicles, while it appears to be sufficient, is not.  The number of vehicles on the road $n$ is sufficiently large to cause the vehicles to interact. Look at a pair of vehicles. If a _lead vehicle_ decelerates to a slower speed than its _following vehicle_ must slow down to avoid a crash. The following driver is risk averse. This interaction moves upstream causing other _leader-follower vehicle pairs_ to slow down as well.  If there are fewer vehicles, the space between vehicles, the _distance headway_ then an individual driver can decelerate  without affecting the vehicle behind it. The vehicles are operating in a _free-flow state_. 

__A Brownian Bridge Model of Speed__

Can we explain this behavior mathematically? Consider the following approach:

First, assume  the speed of a vehicle to be the random variable $U(t)$ and assume the process is defined by a stationary time-series model of the  form: 

* $U(t) =  u + \sigma Z$ 

where $u$ is the assigned speed as specified by the experiment and $\sigma$ is a standard deviation of speed, $Z  \sim N(0,1)$ is a standard normal probability distribution and time $t$ is measured in seconds.  

The next question is how do we assign the model parameters? It is clear that $u$ is the assigned experiment speed but what is its magnitude?  The box plot of speed-density  from our exploratory analyses  showed a connection of speed and density.  The following density-speed plot is a cleaner version of the box plot. 

```{r density-speed,  message = FALSE, fig.height = 5, fig.width = 10, echo = FALSE}
output <- uk_sdk(QUKtdayX)
```

The traffic density $k$ is controlled by the number of vehicles on the road $n$ and ring-road radius $r$:

* $k = n/2\pi r$

Knowing $k$, select $u$ and $\sigma$. Rewrite $U(t)$ as a function of $k$: 

* $U(k,t)  =  u(k) + \sigma(k) * Z$

Assigning $u$ and $\sigma$ are not clear-cut as originally envisioned. In addition to treating speed and speed volatity  as functions of density,  $U(k, t)$ is a _microscale model_ where driver reaction takes place on a time-scale of seconds and  the density-speed plot is a _macroscale model_ derived from 15-minute average data or $\tau$ data.  The scales are incompatable.

Second, the stationary time-series $U(k,t)$ model does not adequately explain driver behavior in a rational manner. Define the time-step equal to $\Delta t = t_i - t_{i-1}$.  Given $Z$ takes on both positive and negative values, a simulation can predict vehicle speeds  flip-flopping back and forth from one second to the next. Obviously, a normal driver does not drive sporatically as described by the model. This flaw is rectified by replacing the $\sigma(k) * Z$ term with a Brownian motion term:

* $U(k,t) = u(k) + \sigma(k)  * \sqrt{\Delta t} * Z$.

Brownian motion or a Wiener process is defined as $W(\Delta t) = W(\Delta t) - W(0)  \sim \sqrt{\Delta t} * Z$ where $\Delta t$ is a time-step. For our application a general formula is needed, $W(\Delta t) \sim N(0,\sigma^2 \Delta t) \sim \sigma * \sqrt{\Delta t} * N(0,1) \sim \sigma * \sqrt{\Delta t} * Z$ where $\sigma(k) = \sigma$ to simplify the notation.

The Brownian motion $U(k,t)$ model eliminates the sporatic flip-flopping described above but it suffers from being a non-stationaey time-series. A driver will now more smoothly accelerate or decelerate. The model is still unacceptable over extended time periods.  For examples, say $\tau$ = 15 minutes,  speeds can reach extraordinary high or low speeds. Negative speeds indicate vehicles are  driving backward, which of course does not represent reality.

This problem is overcome by replacing the Brownian motion $U(k,t)$ model of driver behavior with a __Brownian Bridge model__:

* $U(k,t) = u(k) + \sigma(k) * (W(t) - t/\tau * W(\tau))$.

The following two diagrams compare two randomly chosen simulations for $k$ = 50 and 100 vpm, respectively. They are labelled  "Congested Flow at HCM Capacity" (HCM = Highway Capacity Manual) for $k = 50$ vpm and "Heavily Congested" simulations for $k$ = 100 vpm. The values of $u(k)$ and $\sigma(k)$ are obtained from the density-speed plot. First, we inspect the two simulations.


```{r bridge,  message = FALSE, fig.height = 5, fig.width = 10, echo = FALSE}
par(mfrow = c(1,2))
set.seed(124)
u0 = 41
usd = 11
k0 = 50
output1 <- usdBBridge(41, 11, 90000, 15)
title("","Congested Flow at HCM Capacity")
mtext(bquote(u == .(u0)), side = 3, line = 1)
mtext(bquote(hat(sigma) == .(usd)), side = 3)
mtext(bquote(k == .(k0)), side = 4)
set.seed(120)
output2 <- usdBBridge(18.8, 3.9, 90000, 15)
u0 = 18
usd = 4
k0 = 100
title("","Heavily Congested Flow")
mtext(bquote(u == .(u0)), side = 3, line = 1)
mtext(bquote(hat(sigma) == .(usd)), side = 3)
mtext(bquote(k == .(k0)), side = 4)
```

After time $\tau$ = 15 minutes, the Brownian Bridge model forces a driver in a mathematical sense to return to his or her initial speed $u(k)$, the speed at time $t$ = 0. Between the start and finish times of $t$ = 0 and 15 minutes, the drivers are free to decelerate and accelerate as they wish. The transitions over time are more-or-less smooth as expected. The effect of volatility is in clear evidence.


__Highway Safety__

The Brownian Bridge model described thus far is for a single vehicle. It must be further adapted to simulate: (1) _driver performance in a car-following situation_ and (2) _safety_. These two features are incorporated into the __R cartools__ algorithms. Car-following and safety are strongly associated because drivers are typically risk averse. Drivers avoid crashes.  

Consider a pair of vehicles, one is designated as a _leader_ and the other one designated as a  _follower_. At each time-step in the simulation algorithm, the locations and speeds of the lead and following vehicles and the headway between the two vehicles are calculated. A _safe driving distance_, a distance headway, is defined by the following function:

* $h_{safe} = f(u(k), l_{eff})$, 

where  $u(k)$ is the vehicle speed at density $k$ and $l_{eff}$ is the effective vehicle length. A driver is assumed to keep one vehicle length for each ten mph.  At slow speeds, $h_{safe}$ = two-vehicle lengths or twenty-eight feet headway.  

If there is a safe headway violation, the following vehicle locations and speeds are adjusted to eliminate the violation. If the lead vehicle slows, the following vehicle  must slowdown to avoid a crash if too close. Under a less dense condition, the drivers can vary their speeds without affecting one another. These features are incorporated in __rrtrials__ function, which we denote as:

* $U(t) = U(k, u, \sigma, h_{safe}, t)$.

__Time-Distance Trajectories__

Does __rrtrials__ predict driver behavior that is similar to that shown in the controlled experiment? __rrtrials__  produced the following $t-x$ trajectories. The vehicles are evenly placed along the $x$ axis at $t$ = 0 and the drivers are given the same instructions as the drivers in the Yuki Sugiyama experiment. They are instructed to maintain a constant speed $u$ and headway $h$.   

```{r ringroad1,  message = FALSE, fig.height = 5, fig.width = 10, echo = FALSE}
set.seed(124)
par(mfrow = c(1,2))
k0 <- 50
d <- 22/(pi*k0)
d <- round(d,1)
output1 <- rrtrials(41, 11, 50, 120, 120, 14, 0.75, 10, c(0,120), c(-1000, 7500))
title("","Congested Flow at HCM Capacity")
mtext(bquote(k == .(k0)), side = 4)

h1 <- output1[[5]][1]
hsd1 <- output1[[5]][2]
q1 <- output1[[5]][3]
set.seed(120)
k0 <- 100
d <- 22/(pi*k0)
d <- round(d,1)
output2 <- rrtrials(18.8, 3.9, 100, 120, 120, 14, 0.5, 10, c(0,120), c(-500, 3500))
title("","Heavily Congested Flow")
mtext(bquote(k == .(k0)), side = 4)
h2 <- output2[[5]][1]
hsd2 <- output2[[5]][2]
q2 <- output2[[5]][3]
```

As in the full-scale experiment, the simulated drivers cannot maintain a constant speed or even spacing. After a few seconds, the vehicles begin to interact and form clusters (moving queues) and cannot maintain the constant speeds $u$ is either simulation. 

The volatity effects of $\sigma(k)$ are easily evaluated with visual inspection of the dotted  lines shown in the figures above. The dotted straight line is the trajectory of  the lead vehicle, assuming that its driver could maintain a constant speed $u$ with $\sigma(k)$ = 0. Since $x = u * t$, the slope of the dotted straight line is $u$. Most vehicles in the __rrtrials__ simulation tend to lose speed.

Given these results, __rrtrials__ simulator matches the results observed in the controlled experiment. While there are many variables that are needed to define the results, speed volatility $\sigma$ and the Brownian Bridge $U(k,t)$ are most important.




__Reference__